---
subtitle: "TMA4268 Statistical Learning V2022"
title: "Compulsory exercise 1: Group 16"
author: "Weicheng Hua, Emil Johannese Haugstvedt, Torbjørn Baadsvik"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  # html_document
  pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")

```

```{r,eval=TRUE,echo=FALSE}
# install.packages("knitr") #probably already installed
# install.packages("rmarkdown") #probably already installed
# install.packages("ggplot2") #plotting with ggplot
# install.packages("palmerpenguins")
# install.packages("ggfortify") # For model checking
# install.packages("MASS")
# install.packages("class")
# install.packages("pROC")
# install.packages("plotROC")
# install.packages("boot")
library("knitr")
library("rmarkdown")
library("palmerpenguins")
```

<!--  Etc (load all packages needed). -->

# Problem 1

## a)

The expected MSE on the test set is given by: $$\begin{aligned}
  E[(y_0 - \hat{f}(x_0))^2] &= 
  E[(f(x_0) - \hat{f}(x_0) + \epsilon)^2] \\
  &= E[(f(x_0) - \hat{f}(x_0))^2] + 2E[\epsilon(f(x_0) - \hat{f}(x_0))] + E[\epsilon^2] \\
  &= E[f(x_0)^2 - 2f(x_0)\hat{f}(x_0)] + E[\hat{f}(x_0)^2] + E[\epsilon^2]\\
  &= E[f(x_0)^2 - 2f(x_0)\hat{f}(x_0) + \hat{f}(x_0)^2] 
  + ( E[\hat{f}(x_0)^2] - E[\hat{f}(x_0)]^2) + E[\epsilon^2] \\
  &= E[(f(x_0) - \hat{f}(x_0))^2] + ( E[\hat{f}(x_0)^2] - E[\hat{f}(x_0)]^2) + E[\epsilon^2]\\
  &= E[(f(x_0) - \hat{f}(x_0))^2] + Var[\hat{f}(x_0)] + Var[\epsilon]\\
  &= \text{Squared bias} + \text{Variance of prediction} + \text{Irreducible error}\\ 
\end{aligned}$$

## b)

The squared bias term represents the expected squared deviation between the prediction of the "true" model and the prediction of the fitted model. The variance of prediction term represents the degree to which the prediction of the fitted model can vary depending on the input. Higher variance of prediction means the model can adapt it's prediction to input data to a greater extent than a simpler model, implying that the model is more flexible. However, the increased "adaptability" may be unwanted if it leads to overfitting.

## c)

$$\begin{matrix}
i & ii & iii & iv\\ 
\hline\\
TRUE & FALSE & TRUE & FALSE\\
\end{matrix}$$

## d)

$$\begin{matrix}
i & ii & iii & iv\\ 
\hline\\
TRUE & FALSE & TRUE & FALSE\\
\end{matrix}$$

## e)

```{r}
library(matrixcalc)
mat <- cbind(c(50, 33, 18), c(33, 38, -10), c(18, -10, 72))
is.positive.semi.definite(mat)
```

Answer: iii) 0.76

# Problem 2

Here is a code chunk:

```{r, eval=TRUE}
library(palmerpenguins) # Contains the data set "penguins".
data(penguins)
head(penguins)
```

## a)

## b)

## c)

# Problem 3

Load and prepare data the problem 3

```{r}
# Load libraries
library(tidyverse)
library(class)
library(MASS)
library(palmerpenguins)
library(dplyr)
library(caret)
library(pROC)

# Load penguins data
Penguins <- penguins

# Add binary variable: if adelie
Penguins$adelie <- ifelse(Penguins$species == "Adelie", 1, 0)

# Extract just needed variables and remove na
Penguins_reduced <- Penguins %>% dplyr::select(body_mass_g, flipper_length_mm, adelie) %>% 
  mutate(body_mass_g = as.numeric(body_mass_g),
         flipper_length_mm = as.numeric(flipper_length_mm)) %>% 
  drop_na()

# Set seed for sample extraction
set.seed(4268)

# Find size of training set
training_set_size <- floor(0.7 * nrow(Penguins_reduced))

# Split data randomly into train and test
train_ind <- sample(seq_len(nrow(Penguins_reduced)), size = training_set_size)

train <- Penguins_reduced[train_ind, ]
test <- Penguins_reduced[-train_ind, ]
```

## a)

# i)

```{r}
# Make logistic regression model
model.LR <- glm(adelie ~ body_mass_g + flipper_length_mm,
                 family="binomial",
                 data=train)

# Get classification probabilities
pred.LR.prob <- model.LR %>% predict(test)

# Classify on test set using 0.5 cutoff
pred.LR <- ifelse(pred.LR.prob > 0.5, 1, 0)
```

# ii)

```{r}
# Make quadratic discriminant analysis model
model.QDA <- qda(adelie ~ body_mass_g + flipper_length_mm,
                 data=train)

# Get classification probabilities
pred.QDA.prob<- (model.QDA %>% predict(test))$posterior[0:nrow(test)]

# Classify on test set using 0.5 cutoff
pred.QDA <- ifelse(pred.QDA.prob > 0.5, 1, 0)
```

# iii)

```{r}
# K-nearest neighbor
model.KNN <- knn(train = train,
                 test = test,
                 cl = train$adelie,
                 k = 25,
                 prob = T)

# Get classification probabilities
pred.KNN.prob <- ifelse(model.KNN == 0, 1 - attributes(model.KNN)$prob, attributes(model.KNN)$prob)

# Classify on test set
pred.KNN <- model.KNN[0:nrow(test)]
```

# iv)

```{r}
# Calculate specificity and sensitivity
print("Sensitivity for logistic regression:")
sensitivity(as.factor(pred.LR), as.factor(test$adelie))
print("Sensitivity for QDA:")
sensitivity(as.factor(pred.QDA), as.factor(test$adelie))
print("Sensitivity for KNN:")
sensitivity(as.factor(pred.KNN), as.factor(test$adelie))

print("Specificity for logistic regression:")
specificity(as.factor(pred.LR), as.factor(test$adelie))
print("Specificity for QDA:")
specificity(as.factor(pred.QDA), as.factor(test$adelie))
print("Specificity for KNN:")
specificity(as.factor(pred.KNN), as.factor(test$adelie))

```

## b)

# i)

```{r}
# ROC for different classifiers
roc.LR <- roc(test$adelie, pred.LR.prob)
roc.KNN <- roc(test$adelie, pred.KNN.prob)
roc.QDA <- roc(test$adelie, pred.QDA.prob)

roc.list <- list(
  "Logistic regression" = roc.LR,
  "KNN" = roc.KNN,
  "QDA" = roc.QDA)

# Plot ROC
ggroc(roc.list, aes = "col",legacy.axes = TRUE) + 
  geom_abline() + 
  theme_classic() +
  ggtitle("ROC") +
  labs(x = "1 - Specificity",
       y = "Sensitivity",
       col = "Model type")

# AUC for the classifiers
roc.LR$auc
roc.KNN$auc
roc.QDA$auc
```

# ii)

The ROC curve shows how the binary classifiers are performing with a varying threshold. The plot shows the true positive rate, or sensitivity, against the true negative rate, or 1 - specificity. If a classifier is on the black diagonal line it performs just as good as guessing randomly and if it lays above the diagonal line it performs better. This means that a classifier is better the further up in the left corner its ROC curve are. In this case the logistic regression and the QDA performs better than the KNN classifier.

The AUC, or area under the curve, is just the area under the ROC curve and provides an aggregate measure of all classifiers for all thresholds. As expected, the AUC for the logistic regression and the QDA are higher than the one for the KNN indicating better performance.

# iii)

We would say the logistic regression model is the most interpretable model. This model provides easy to understand betas telling about the realtionship between the body mass and flipper length and if the penguin is Adelie or not. The two other models does not provide the same information about the data.

## c)

```{r}
model.LR$coefficients
```

Increasing the body mass of penguin with 1000g will result in the following change in the odds

\$\$ \begin{equation}

exp(0.0007120 * 1000) = 2.038

\end{equation} \$\$

The odds increases by a factor of 2.038. Alternative iii.

## d)

```{r}
# Get all data, from both test and training
penguins.pred.all <- Penguins_reduced %>% dplyr::select(-adelie)

# Classify on both test and training data using LR model
pred.all <- ifelse(model.LR %>% predict(penguins.pred.all) > 0.5, 1, 0)

penguins.pred.all <- Penguins_reduced %>% dplyr::mutate(pred = pred.all)

# Plot result
ggplot(penguins.pred.all, aes(x=flipper_length_mm, y=body_mass_g)) + 
  ggtitle("Classification of penguins") +
  geom_point(aes(col = as.factor(adelie), shape = as.factor(pred))) + 
  labs(x = "Flipper length [mm]",
       y = "Body mass [g]",
       col = "True class",
       shape = "Predicted class") + 
  scale_colour_discrete(labels=c("Not Adelie", "Adelie")) +
  scale_shape_discrete(labels=c("Not Adelie", "Adelie"))
```

# Problem 4
